Predicting the effects of character displacement on food-web dynamics

### 2019-11-18 manuscript_ECD_model.Rmd
incorporate final (minor) edits to main and supplementary text.

---
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    citation_package: natbib 
header-includes:
  - \usepackage{fullpage}
  - \linespread{1.5} 
  - \usepackage{lineno}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
bibliography: ReferencesAbbrev
biblio-style: EcolLett
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
citecolor: black
---

@@

*Data accessibility*: No new data were used. All code used to reproduce this manuscript and simulations within are publically available on GitHub (https://github.com/mabarbour/ECD_model) and have been archived with Zenodo (INSERT DOI).

@@

*Running title*: Character divergence destabilizes food webs

@@

<!-- Remove rest of title page for bioRxiv submission -->
\bigskip

@@

Ecological character displacement is an adaptive process that generally increases phenotypic diversity. 
Despite the fact that this diversification is due to an eco-evolutionary feedback between consumers competing for shared resources, its consequences for food-web dynamics have not been examined.
Here, I study a model of two consumers competing for two shared resources to examine how character displacement in consumer attack rates affects resource abundances and the resilience of food webs to perturbations.
I found that character displacement always strengthened consumer-resource interactions whenever consumers competed for resources that occurred in different habitats.
This increase in interaction strength resulted in lower resource abundances and less resilient food webs.
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
Taken together, my results show that the adaptive process of character displacement may come with the ecological cost of decreasing food-web resilience.

@@

Although models that included resources led to insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics has received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of coexistence theory [@Lawlor1976; @Germain2018; @Bassar2017; @McPeek2019]. 
For example, early theoretical work showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition [@Lawlor1976]. 
Yet, this reduction in interspecific competition may, at the same time, increase interspecific interactions between specialized consumers and their resources. 
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often suppresses the abundance of resources, which if sufficient enough, can generate oscillations and less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that examines how ecological character displacement affects consumer-resource dynamics in a food-web context. 
I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? 
To test the generality of these effects, I explored different ecological foraging scenarios and evolutionary tradeoffs in consumer attack rates. 
I found that the adaptive process of character displacement often comes with an ecological cost; resulting in food webs with lower resource availability and that are less resilient to perturbations.

@@

## Underlying Consumer-Resource Dynamics

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ij}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ij}(R_i)$ represents consumer $j$'s feeding rate on resource $i$ (i.e. functional response). 
This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates in lakes) rather than a scenario where resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). 
Importantly, inferences about ecological character displacement can only be made by comparing food webs with and without a competing consumer [@Schluter1992].
Therefore, I arbitrarily set $C_2=0$ to create a food-web without a competing consumer for these comparisons.

## Foraging Scenarios

I studied three different foraging scenarios. 
In the first, I assume that consumers can forage for both resources simultaneously (fig. \ref{fig:foraging_scenarios} top) and their feeding rates increase linearly with resource abundance, such that:

@@

where $a_{ij}$ is the attack rate of consumer $j$ on resource $i$. 
This first scenario is the starting point for many models of resource competition [@MacArthur1972]; however, it does not reflect many food webs where consumers are mobile and their foraging behavior links resources that occur in different habitats [@McCann2005]. 
The second scenario accounts for this spatial context (fig. \ref{fig:foraging_scenarios} bottom) and takes the form:

@@

where $w_{ij}$ represents the proportion of time consumer $j$ spends foraging in a habitat where only resource $i$ is found (i.e. habitat preference). 
Note that since $w_{ij}$ is a proportion that $w_{1,j}=1-w_{2,j}$. <!-- $\sum_jw_{ij}=1$.<!-- $w_{ij}=1-w_{ii}$. -->
Finally, it is well known that consumer feeding rates often saturate at high resource abundances [@Holling1959; @Rosenzweig1963; @Murdoch2003; @McCann2011] and that consumers do not usually spend a fixed proportion of time in a particular habitat [@McCann2005]. 
The third scenario accounts for these biological realities and takes the form [derived by @McCann2005]:     
\begin{equation} \label{eq:4}
  F_{ij}(R_i)=\frac{a_{ij}W_{ij}R_i}{1+a_{1,j}h_{1,j}W_{1,j}R_1+a_{2,j}h_{2,j}W_{2,j}R_2}
\end{equation}

where consumer $j$'s feeding rate on resource $i$ is influenced by the abundance of each resource; saturates as resource abundances increase (due to handling times $h_{ij}$); and consumer habitat preferences are modified by the relative abundance of resources, such that: $W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}$.

Previous studies have analyzed the evolution of consumer attack rates in the first two foraging scenarios using an Adaptive Dynamics approach, with the general result being character divergence [@Lawlor1976; @Abrams1986]. 
I also used an Adaptive Dynamics approach to analyze character displacement in the third foraging scenario, and I too observed character divergence (detailed analysis in Appendix S1 of Supplementary Information).
I say consumers have undergone character divergence if their evolved attack rates are more specialized when evolving with vs. without a competing consumer. 
Specialization of consumer $j$ on resource $1$ is measured as $\frac{a_{1,j}}{a_{1,j}+a_{2,j}}$, where a value of 0.5 is a complete generalist ($a_{1,j}=a_{2,j}$), and a value of 1 is a complete specialist ($a_{2,j}=0$).
Values less than 0.5 indicate specialization on the other resource.

```{r foraging_scenarios, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:foraging_scenarios}**Ecological foraging scenarios.** I examined whether the effect of ecological character displacement on food-web dynamics depended on whether consumers competed for resources that occurred in the same vs. different habitats. The width of each arrow corresponds to initial attack rates (*a*~ij~) of consumer *j* on resource *i*. Note that *C*~1~ was pre-adapted to *R*~1~ (*a*~11~ > *a*~21~), while *C*~2~ was a mirror image, being pre-adapted to *R~2~* (*a*~22~ > *a*~12~). In each scenario, I assumed consumer feeding rates increased linearly with resource abundance. I also relax this assumption and consider a more realistic functional response when resources occurred in different habitats."}
knitr::include_graphics("Fig_1_ForagingScenarios")
```

## Food-web Dynamics

Given that character divergence occurred across these foraging scenarios, I focus here on its consequences for food-web dynamics.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when the rates of change in equation \ref{eq:1} are 0, and solving the system at this point gives equilibrium abundances for each resource ($\hat R_i$) and each consumer ($\hat C_j$). 
I also compared the local stability of these food webs using standard methods [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda_{\text{max}}$, of the matrix of partial derivatives of all differential equations with respect to all variables evaluated at equilibrium. 
If $-\lambda_{max}>0$, then the food web will return to equilibrium after a small perturbation (i.e. locally stable), with more positive values indicating a faster return time.
If $-\lambda_{max}<0$, then the food web is not locally stable.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_i$ and $K=K_i$) as well as consumers ($e=e_{ij}$; $h=h_{ij}$; $m=m_j$), except that consumer attack rates and their habitat preferences (if present) are mirror images of each other ($a_{11}=a_{22}$; $a_{12}=a_{21}$; $w_{11}=w_{22}$). 
Note that I arbitrarily set $C_1$ as being pre-adapted to $R_1$ ($a_{11} > a_{21}$; $w_{11} > 0.5$), and therefore $C_2$ was pre-adapted to $R_2$.
Controlling for other sources of variability allowed me to isolate the general effects of character divergence.
All mathematical derivations were conducted in Mathematica [@Mathematica] and are provided in the Supplementary Information (Appendix S1-3).

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
Specifically, after letting consumer and resource abundances reach a steady state, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ are higher than the extremes (when $n>1$, green line in fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ have the same total attack rate (blue line in fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource abundances to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}$ is constrained to a maximum of 1 and minimum of 0). 
Unless otherwise noted, I conducted simulations with the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2; *h* = 0.4; and $w_{11} = w_{22} = 0.6$. 
I set an initial value of $a_{11} = a_{22} = 1.2$, while $a_{12}$ and $a_{21}$ depended on the value of *n*. 
I set initial consumer and resource abundances to: $R_1 = R_2 = 2$; $C_1 = C_2 = 1$.
All simulations were conducted in R [@R] and the code to reproduce these simulations is publically available on GitHub (https://github.com/mabarbour/ECD_model) and have been archived with Zenodo (INSERT DOI).

@@

```{r tradeoff, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:tradeoff}**Evolutionary tradeoffs in consumer attack rates.** In each foraging scenario, I explored the effects of three different tradeoffs: intermediate combinations of attack rates (*a*~1,*j*~, *a*~2,*j*~) are higher than the extremes (green line, *n* > 1); extreme combinations of attack rates are higher than intermediate investments (orange line, *n* < 1); and all combinations of attack rates have the same total attack rate (blue line, *n* = 1). Points corresponding to attack rates at the beginning of the simulation for *C*~1~, which was pre-adapted to *R*~1~ (*a*~11~ > *a*~12~)."}
knitr::include_graphics("Fig_2_Tradeoffs.pdf")
```

@@

## Resources occur in same habitat

@@

I found that the shape of the tradeoff function qualitatively affects the relationship between character displacement and resource abundances in this scenario (fig. \ref{fig:plot_fig3} middle left).
For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (fig. \ref{fig:plot_fig3} top left) and character displacement has no effect on resource abundances (fig. \ref{fig:plot_fig3} middle left). 
If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (fig. \ref{fig:plot_fig3} middle left). 
This is because the total attack rate of consumers is maximized at intermediate values ($a_{1,j}=a_{2,j}$) and decreases as consumers diverge (fig. \ref{fig:plot_fig3} top left). 
When the tradeoff is concave up (orange lines), character displacement suppresses resource abundances due to the increase in total attack rates (fig. \ref{fig:plot_fig3} top left). 
Although the equation I derived for resource abundances was for the scenario where both consumers and both resources were present, it accurately predicts the abundance of resources when a single consumer reaches its evolutionary stable strategy (ESS; triangles on respective colored lines in fig. \ref{fig:plot_fig3} middle left). 
This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in fig. \ref{fig:plot_fig3} left panels), resulting in equivalent resource abundances.

The effect of character displacement on resources corresponds to its impact on food-web stability. 
For example, when character displacement decreases resource abundances (orange points in fig. \ref{fig:plot_fig3} middle left), there is a decrease in food-web stability (fig. \ref{fig:plot_fig3} bottom left). 
Character divergence may not affect or even increase food-web stability (blue and green lines in fig. \ref{fig:plot_fig3} bottom left); however, evolution does not favor strong divergence in these scenarios (blue and green points in fig. \ref{fig:plot_fig3}), which dampens these contingent effects.
Note that the dip in stability occurs when both consumers evolve to be generalists, a situation that is not favored in any of the foraging scenarios we examined (fig. \ref{fig:plot_fig3} bottom).

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{1,j}>0.5$ and $a_{1,j}>a_{2,j}$), then the effective attack rate of consumers ($w_{1,j}a_{1,j}+w_{2,j}a_{2,j}$) will always increase, regardless of the tradeoff (fig. \ref{fig:plot_fig3} top right). 
Thus, character divergence always results in resource suppression (fig. \ref{fig:plot_fig3} middle right). 
Note that the shape of the tradeoff can modify the effect of character displacement. 
This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached an ESS (triangles in fig. \ref{fig:plot_fig3} middle right). 
In contrast, resource abundances reach a similar value when consumers evolve in the presence of a competitor (circles in fig. \ref{fig:plot_fig3} middle right), because character displacement tends to reach a constraint of complete specialization. 
It is worth noting that resource abundances are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in fig. \ref{fig:plot_fig3} middle right).
This is because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in fig. \ref{fig:plot_fig3} right panels).

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (fig. \ref{fig:plot_fig3}). 
Specifically, character divergence decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (fig. \ref{fig:plot_fig3}). 
For example, when the tradeoff is concave up (orange), the initial two-consumer food web (small circle) is more stable than when there is only one consumer (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).

```{r Fig 3, include=FALSE}
Macs <- plot_grid(
  Mac_attack + xlab("") + ylab(bquote(atop("Effective attack rate","("*a[1*","*j]+a[2*","*j]*")"))) + ggtitle("Resources: Same Habitat\n") + theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), plot.title = element_text(size = 10), axis.title.y=element_text(vjust=0, size=12)), 
  Mac_resources + xlab("") + ylab(bquote(atop("Resource abundance","("*R[1]+R[2]*")"))) + theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), axis.title.y=element_text(vjust=0, size=12)), 
  Stability_Mac + scale_x_continuous(labels = c(0,0.25,0.5,0.75,1)) + xlab("") + ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) + theme(legend.position = "none", axis.text = element_text(size=10), axis.title.y=element_text(vjust=0, size=12)), 
  ncol=1, align='v'
)

>>>>

---
output:
  pdf_document:
    keep_tex: true
    toc: false
    number_sections: false
    df_print: paged
    fig_caption: true
    citation_package: natbib 
header-includes:
  - \usepackage{fullpage}
  - \linespread{1.5} 
  - \usepackage{lineno}
  - \usepackage{caption,setspace}
  - \captionsetup[figure]{font=small}
bibliography: ReferencesAbbrev
biblio-style: EcolLett
fontsize: 11pt
fontfamily: mathpazo
documentclass: article
linkcolor: black
urlcolor: black
citecolor: black
---

@@

*Data accessibility*: No new data were used. All code used to reproduce this manuscript and simulations within are publically available on GitHub (https://github.com/mabarbour/ECD_model)<!-- and have been archived with Zenodo (INSERT DOI)-->.

@@

*Running title*: Character displacement destabilizes food webs

@@

<!-- Remove rest of title page for bioRxiv submission
\bigskip

@@

Ecological character displacement is an adaptive process that generally increases phenotypic diversity. Despite the fact that this diversification is due to an eco-evolutionary feedback between consumers competing for shared resources, its consequences for food-web dynamics have received little attention. Here, I study a model of two consumers competing for two shared resources to examine how character displacement in consumer attack rates affects resource abundances and the resilience of food webs to perturbations. I found that character displacement always strengthened consumer-resource interactions whenever consumers competed for resources that occurred in different habitats. This increase in interaction strength resulted in lower resource abundances and less resilient food webs. This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios. Taken together, my results show that the adaptive process of character displacement may come with the ecological cost of decreasing food-web resilience.

@@

Although models that included resources led to insights about the evolution of character displacement, the ecological feedback onto consumer-resource dynamics has received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of coexistence theory [@Lawlor1976; @Germain2018; @Bassar2017; @McPeek2019]. 
For example, early theoretical work showed that character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition [@Lawlor1976]. 
Yet, this reduction in interspecific competition may, at the same time, increase interspecific interactions between specialized consumers and their resources. 
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often suppresses the abundance of resources, which if sufficient enough, can generate oscillations and less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that examines how ecological character displacement affects consumer-resource dynamics in a food-web context.
Specifically, I sought to answer the question: how does character displacement in consumer attack rates affect resource abundances and food-web stability?
To test the generality of these effects, I explored different ecological foraging scenarios and evolutionary tradeoffs in consumer attack rates. 
I found that the adaptive process of character displacement often comes with an ecological cost; resulting in food webs with lower resource availability and that are less resilient to perturbations.

@@

## Underlying consumer-resource dynamics

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ij}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ij}(R_i)$ represents consumer $j$'s feeding rate on resource $i$ (i.e., its functional response). 
This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates in lakes) rather than a scenario where resources are better characterized by a continuous trait distribution (e.g., seed size; see @Taper1985 for an example). 
Importantly, inferences about character displacement can only be made by comparing food webs with and without a competing consumer [@Schluter1992].
Therefore, I arbitrarily set $C_2=0$ to create a food-web without a competing consumer for these comparisons.

## Foraging scenarios

I studied three different foraging scenarios. 
In the first, I assume that consumers can forage for both resources simultaneously (Fig. \ref{fig:foraging_scenarios}a) and their feeding rate increases linearly with resource abundance, such that:

@@

where $a_{ij}$ is the attack rate of consumer $j$ on resource $i$. 
This first scenario is the starting point for many models of resource competition [@MacArthur1972]; however, it does not reflect many food webs where consumers are mobile and their foraging behavior links resources that occur in different habitats [@McCann2005]. 
The second scenario accounts for this spatial context (Fig. \ref{fig:foraging_scenarios}b) and takes the form:

@@

where $w_{ij}$ represents the proportion of time consumer $j$ spends foraging in a habitat where only resource $i$ is found (i.e., its habitat preference). 
Note that since $w_{ij}$ is a proportion that $w_{1,j}=1-w_{2,j}$. <!-- $\sum_jw_{ij}=1$.<!-- $w_{ij}=1-w_{ii}$. -->
Finally, it is well known that consumer feeding rates often saturate at high resource abundances [@Holling1959; @Rosenzweig1963; @Murdoch2003; @McCann2011] and that consumers do not usually spend a fixed proportion of time in a particular habitat [@McCann2005]. 
The third scenario accounts for these biological realities and takes the form [derived by @McCann2005]:     
\begin{equation} \label{eq:4}
  F_{ij}(R_i)=\frac{a_{ij}W_{ij}R_i}{1+a_{1,j}h_{1,j}W_{1,j}R_1+a_{2,j}h_{2,j}W_{2,j}R_2}
\end{equation}

where consumer $j$'s feeding rate on resource $i$ is influenced by the abundance of each resource; saturates as resource abundances increase (due to handling time $h_{ij}$); and consumer habitat preferences are modified by the relative abundance of resources, such that: $W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}$.

Previous studies have analyzed the evolution of consumer attack rates in the first two foraging scenarios using an Adaptive Dynamics approach, with the general result being divergent character displacement [@Lawlor1976; @Abrams1986]. 
I also used an Adaptive Dynamics approach to analyze the evolution of consumer attack rates in the third foraging scenario, and I too observed divergent character displacement (detailed analysis given in Appendix S1).
I say consumers have undergone divergent character displacement if their evolved attack rates are more specialized when evolving with vs. without a competing consumer. 
Specialization of consumer $j$ on resource $1$ is measured as $\frac{a_{1,j}}{a_{1,j}+a_{2,j}}$, where a value of 0.5 is a complete generalist ($a_{1,j}=a_{2,j}$), and a value of 1 is a complete specialist ($a_{2,j}=0$).
Values less than 0.5 indicate specialization on the other resource. 
Since I did not observe convergent character displacement in any of the foraging scenarios I analyzed, I refer to divergent character displacement as simply (ecological) character displacement throughout the rest of the text.

```{r foraging_scenarios, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:foraging_scenarios}**Ecological foraging scenarios.** I examined whether the effect of ecological character displacement on food-web dynamics depended on whether consumers competed for resources that occurred in the same (a) vs. different habitats (b). Note that inferences about character displacement can only be made by comparing food webs with (right) and without (left) a competing consumer, so I arbitrarily set *C*~2~ = 0 for these comparisons. The width of each arrow corresponds to the initial attack rate (*a*~ij~) of consumer *j* on resource *i*. Note that *C*~1~ was pre-adapted to *R*~1~ (*a*~11~ > *a*~21~), while *C*~2~ was a mirror image, being pre-adapted to *R~2~* (*a*~22~ = *a*~11~; *a*~12~ = *a*~21~). In each scenario, I assumed consumer feeding rates increased linearly with resource abundance. I also relax this assumption and consider a more realistic functional response when resources occurred in different habitats (b)."}
knitr::include_graphics("Fig_1_ForagingScenarios")
```

## Food-web dynamics

Given that character displacement occurred across these foraging scenarios, I focus here on its consequences for food-web dynamics.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when there is no change in the population growth rates of consumers and resources (i.e., the rates of change in equation \ref{eq:1} are 0), and solving the system at this point gives equilibrium abundances for each resource ($\hat R_i$) and each consumer ($\hat C_j$). 
I also compared the local stability of these food webs using standard methods [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda_{\text{max}}$, of the matrix of partial derivatives of each species' population growth rate (given by equation \ref{eq:1}) with respect to each species' abundance evaluated at equilibrium. 
If $-\lambda_{max}>0$, then the food web will return to equilibrium after a small perturbation (i.e., it is locally stable), with more positive values indicating a faster return time.
If $-\lambda_{max}<0$, then the food web is not locally stable.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_i$ and $K=K_i$) as well as consumers ($e=e_{ij}$; $h=h_{ij}$; $m=m_j$), except that consumer attack rates and their habitat preferences (if present) are mirror images of each other ($a_{11}=a_{22}$; $a_{12}=a_{21}$; $w_{11}=w_{22}$). 
Note that I arbitrarily set $C_1$ as being pre-adapted to $R_1$ ($a_{11} > a_{21}$; $w_{11} > 0.5$), and therefore $C_2$ was pre-adapted to $R_2$.
Controlling for other sources of variability allowed me to isolate the general effects of character displacement.
All mathematical derivations were conducted in Mathematica [@Mathematica] and are provided in Appendices S1-3.

To gain insight to the eco-evolutionary feedback generated by character displacement, I conducted simulations using an Adaptive Dynamics approach. 
Specifically, after letting consumer and resource abundances reach a steady state, I created a mutant consumer by randomly choosing one and modifying its attack rate on one resource by either subtracting or adding a small constant (0.01 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ are higher than the extremes (when $n>1$, green line in Fig. \ref{fig:tradeoff}) or, conversely, where the two extremes are higher than intermediate investments (when $n<1$, orange line in Fig. \ref{fig:tradeoff}). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{\text{1},j}$ and $a_{\text{2},j}$ have the same total attack rate (blue line in Fig. \ref{fig:tradeoff}). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource abundances to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g., $\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}$ is constrained to a maximum of 1 and minimum of 0). 
Unless otherwise noted, I conducted simulations with the following parameter values: *r* = 1; *K* = 4; *e* = 0.8; *m* = 1; *A* = 2; *h* = 0.4; and $w_{11} = w_{22} = 0.6$. 
I set an initial value of $a_{11} = a_{22} = 1.2$, while $a_{12}$ and $a_{21}$ depended on the value of *n*. 
I set initial consumer and resource abundances to: $R_1 = R_2 = 2$; $C_1 = C_2 = 1$.
All simulations were conducted in R [@R] and the code to reproduce these simulations is publically available on GitHub (https://github.com/mabarbour/ECD_model)<!-- and has been archived with Zenodo (INSERT DOI)-->.

@@

```{r tradeoff, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:tradeoff}**Evolutionary tradeoffs in consumer attack rates.** In each foraging scenario, I explored the effects of three different tradeoffs: intermediate combinations of attack rates (*a*~1,*j*~, *a*~2,*j*~) are higher than the extremes (green line, *n* > 1); extreme combinations of attack rates are higher than intermediate investments (orange line, *n* < 1); and all combinations of attack rates have the same total attack rate (blue line, *n* = 1). Points corresponding to attack rates at the beginning of the simulation for *C*~1~, which was pre-adapted to *R*~1~ (*a*~11~ > *a*~12~). Note that *C*~2~ was a mirror image of *C*~1~, being pre-adapted to *R*~2~ (*a*~22~ = *a*~11~; *a*~12~ = *a*~21~)."}
knitr::include_graphics("Fig_2_Tradeoffs.pdf")
```

@@

## Resources occur in the same habitat

@@

I found that the shape of the tradeoff function qualitatively affects the relationship between character displacement and resource abundances in this scenario (Fig. \ref{fig:plot_fig3}a,b).
For example, if consumer's are constrained by a linear tradeoff (blue lines), then there is no net change in total attack rate (Fig. \ref{fig:plot_fig3}a) and character displacement has no effect on resource abundances (Fig. \ref{fig:plot_fig3}b). 
If the tradeoff is concave down (green lines), then resource abundances can actually increase under character displacement (Fig. \ref{fig:plot_fig3}b). 
This is because the total attack rate of consumers is maximized at intermediate values ($a_{1,j}=a_{2,j}$) and decreases as consumers diverge (Fig. \ref{fig:plot_fig3}a). 
When the tradeoff is concave up (orange lines), character displacement suppresses resource abundances due to the increase in total attack rates (Fig. \ref{fig:plot_fig3}a,b). 
Although the equation I derived for resource abundances was for the scenario where both consumers and both resources were present, it accurately predicts the abundance of resources when a single consumer reaches its evolutionary stable strategy (ESS; triangles on respective colored lines in Fig. \ref{fig:plot_fig3}b). 
This is because a single consumer evolves to be a generalist that has equal attack rates on each resource (triangles at 0.5 along x-axis in Fig. \ref{fig:plot_fig3}a), resulting in equivalent resource abundances.

The effect of character displacement on resources corresponds to its impact on food-web stability. 
For example, when character displacement decreases resource abundances (orange points in Fig. \ref{fig:plot_fig3}b), there is also a decrease in food-web stability (Fig. \ref{fig:plot_fig3}c). 
Character displacement may not affect or even increase food-web stability (blue and green lines in Fig. \ref{fig:plot_fig3}c); however, evolution does not favor strong divergence in these scenarios (blue and green points in Fig. \ref{fig:plot_fig3}a), which dampens these contingent effects.
Note that the dip in stability occurs when both consumers evolve to be generalists, a situation that is not favored in any of the foraging scenarios I examined (Fig. \ref{fig:plot_fig3}c).

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g., $w_{1,j}>0.5$ and $a_{1,j}>a_{2,j}$), then the effective attack rate of consumers ($w_{1,j}a_{1,j}+w_{2,j}a_{2,j}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:plot_fig3}d). 
Thus, character displacement always results in resource suppression (Fig. \ref{fig:plot_fig3}e). 
Note that the shape of the tradeoff can modify the effect of character displacement. 
This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached an ESS (triangles in Fig. \ref{fig:plot_fig3}e). 
In contrast, resource abundances reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:plot_fig3}e), because character displacement tends to reach a constraint of complete specialization. 
It is worth noting that resource abundances are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:plot_fig3}e).
This is because consumers actually evolve slightly specialized attack rates on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:plot_fig3}d).

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (Fig. \ref{fig:plot_fig3}f). 
Specifically, character displacement decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:plot_fig3}c). 
For example, when the tradeoff is concave up (orange), the initial two-consumer food web (small circle) is more stable than when there is only one consumer (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).

```{r Fig 3, include=FALSE}
Macs <- plot_grid(
  Mac_attack + xlab("") + ylab(bquote(atop("Effective attack rate","("*a[1*","*j]+a[2*","*j]*")"))) +
    theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), plot.title = element_text(size = 10), axis.title.y=element_text(vjust=0, size=12)), 
  Mac_resources + xlab("") + ylab(bquote(atop("Resource abundance","("*R[1]+R[2]*")"))) + theme(legend.position = "none", axis.text = element_text(size=10), axis.text.x = element_blank(), axis.title.y=element_text(vjust=0, size=12)), 
  Stability_Mac + scale_x_continuous(labels = c(0,0.25,0.5,0.75,1)) + xlab("") + ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) + theme(legend.position = "none", axis.text = element_text(size=10), axis.title.y=element_text(vjust=0, size=12)), 
  ncol=1, align='v', labels = c("","b)","c)"), hjust = -4, vjust = 0.5, label_size = 10
)
Macs_title <- ggdraw() + draw_label("Resources: Same Habitat", fontface = 'bold', size = 10, hjust = 0.2)
plot_Macs <- plot_grid(Macs_title, Macs, ncol = 1, rel_heights = c(0.1, 1))
